Long term cloud enhancement events of global solar irradiance over Thessaloniki, Greece

### 2024-06-12 article/article.tex
Commit 2024-06-12 19:11

We studied the enhancement of the global horizontal irradiance (GHI),
over Thessaloniki, for the period 1993 -- 2023, on one minute average
records. We identify the cloud enhancement events (CE), by creating an
appropriate clear sky irradiance reference with the use of a radiation
transfer model, and aerosol optical depth (AOD) data, from the AERONET,
and a collocated Brewer photospectrometer. We found that there is a
trend of CE cases of \(45.6\pm 21.9\,\text{cases}/\text{year}\), and a
mean total energy change, of the CE events by
\(116.9\pm 67.8\,\text{kj}/\text{year}\) with the peak of the CE events
are observed during May and June. An analysis of the total duration of
CE events, showed that durations longer than 5 minutes are very rare,
with exceptions lasting over an hour. We have detected enhancements over
the total solar irradiance, without exceeding the maximum values
reported for other sites, with more favorable CE conditions.

@@

For the detection of CE cases we established a baseline of irradiance
above which we characterized each data point as an enhancement event and
calculated the over irradiance (OIR). The OIR is defined as the
irradiance difference of the measured one-minute \(\text{GHI}_i\) from
the \(\text{GHI}_\text{ref}\) corresponding to cloud-free atmosphere.
First, we tested two simple approaches for the determination of
\(\text{GHI}_\text{ref}\): The Haurwitz's model \citep{Haurwitz1945},
which is a simple clear sky model and was already adjusted and applied
to our data \citep{Natsis2023}, and the total solar irradiance (TSI) at
the top of the atmosphere, adjusted for the Sun-Earth distance. We have
tested both methods by using an appropriate relative threshold and/or an
additional constant offset. The initial results showed that we can
detect a big portion of the CE events. However, by inspecting the daily
plots of irradiance it became evident that changes in atmospheric
conditions introduced numerous false positive and false negative
results. The main reason for these discrepancies is the variability of
the effects of aerosols and water vapor which were not taken into
account in the two simple methods. To produce a more representative
reference we included the effects of these factors through variables
using a radiative transfer model (RTM). The applied methodology is
discussed in section\nobreakspace{}\ref{rtmcs}

@@

The above clear sky reference values are based on the climatological AOD
and WC, hence they cannot describe accurately the long term variation of
\(\text{GHI}_\text{ref}\) due to changes in the two atmospheric
constituents, mainly AOD. As reported by \citet{Natsis2023}, there is a
long term brightening effect in the GHI data of Thessaloniki for the
period 1993 -- 2023, which for clear-sky data was attributed to
long-term changes in aerosol effects. Therefore, an adjustment of the
\(\text{GHI}_\text{ref}\) used in this study is necessary. As AERONET
data start only in 2003, we used for the period 1993 -\/-- 2005
estimates of changes in AOD at \(340\,\text{nm}\) derived from a
collocated Brewer spectrophotometer derived from two sources. For 1993
-- 2005 \citep{Kazadzis2007} to calculate the trend in
\(\text{GHI}_\text{ref}\) during this period.

@@

Following the application of the above discussed methodology to the
entire dataset (6.1 million of one-minute GHI measurements),
\(1.764\,\%\) were identified as CE events and \(0.036\,\%\) as ECE
events. The highest recorded GHI due to CE was
\(1416.6\,\text{W}/\text{m}^2\) on 24~May 2007, corresponding to OIR of
\(345.9\,\text{W}/\text{m}^2\). The stronger CE event of \(49.7\,\%\)
above the clear sky threshold was observed on 28~October 2016. In the
following sections we are discussing the long-term trends and
variability of the CE events as well as of the corresponding OIR and
excess irradiation.

@@

Another interesting aspect of the CE events is their distribution within
the year. Figure\nobreakspace{}\ref{fig:relative-month-occurrences}
shows the monthly box and whisker plot of the CE number of occurrence
normalized with the highest median value, that occurs in June, depicting
a clear seasonal cycle. Although CE events are present throughout the
year, the most active period is during May and June. During the winter
(December -- February), the number of CE cases is about \(25\,\%\) of
the maximum, while in the intermediate months, the number of occurrences
gradually ramps between the maximum and minimum. This seasonality is a
combined effect of different factors, among them the types of clouds,
their frequency of occurrence the seasonally varying relative position
of the sun, as well as the local landscape characteristics that may
influence the formation of clouds. Unfortunately, lack of detailed data
on cloud formation, type and location is not allowing further analysis.
The interannual variability of the monthly CE events is quite high as
manifested by the large monthly extremes, especially in the summer.

@@

\caption{Seasonal variability of the number of CE events normalized to the maximum occurring in June, in the form of a box and whisker plot. The monthly values have been also normalized to the relative abundance of valid GHI observations. The box contains the data between the lower $25\,\%$ and the upper $75\,\%$ percentiles,The thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values and the solid circles are outliers.}\label{fig:relative-month-occurrences}
\end{figure}

The distribution of the number of CE events as a function of OIR is
shown in Figure\nobreakspace{}\ref{fig:ovir-distribution}. Apparently,
there is an inverse relation between the frequency of CE events and OIR
with an exponential-like decline. This is expected, as the stronger the
CE events become, the rarer are the conditions favoring the occurrence
of CE events. For the majority (over \(62\,\%\)) of the CE events the
OIR is below the long term average of \(39.9\,\text{W}/\text{m}^2\),
while about \(8.1\,\%\) of the events correspond to OIR larger than
\(100\,\text{W}/\text{m}^2\) and up to highest value of
\(412.4\,\text{W}/\text{m}^2\). This distribution is indicative of the
magnitude and the probability of the expected CE events over
Thessaloniki. Similar distribution of CE events, albeit with larger OIR
values, has been reported by \citet{Vamvakas2020}, for the city of
Patras.This site is located \(2.5^\circ\) south of Thessaloniki and is
exposed to air masses coming mainly from the eastern Mediterranean
resulting in different cloud patterns, that may affect the
characteristics and magnitude of the CE events.

@@

The relation between the duration and the mean OIR intensity of the
groups have also been studied (Figure\nobreakspace{}\ref{fig:group-2d}).
We observed that the GCE events tend to have either long duration, or
large intensity. Events with strong enhancement and long duration are
very rare. Similar results on this relation, have been reported by
\citet{Zhang2018}, on a study using a far higher sampling rate than
ours.

@@

By creating a clear sky approximation of the GHI, which represents the
long- and short-term variation of the expected clear sky GHI, we were
able to identify cases of CE events. After analyzing the CE cases, we
found an increase of \(45.6\pm 21.9\,\text{cases}/\text{year}\), with
the mean annual total energy of the CE events increasing with a rate of
\(116.9\pm 67.8\,\text{kj}/\text{year}\). The most active season of CE
events over Thessaloniki is concentrated on early summer, on May and
June.

>>>>

We studied the enhancement of the global horizontal irradiance (GHI),
over Thessaloniki, for the period 1993 -- 2023, on one minute average
records. We identify the cloud enhancement events (CE), by creating an
appropriate clear sky irradiance reference with the use of a radiation
transfer model, and aerosol optical depth (AOD) data, from the AERONET,
and a collocated Brewer photospectrometer. We found that there is a
trend of CE cases of \(45.6\pm 21.9\,\text{cases}/\text{year}\), and a
mean total energy change, of the CE events by
\(116.9\pm 67.8\,\text{kJ}/\text{year}\) with the peak of the CE events
are observed during May and June. An analysis of the total duration of
CE events, showed that durations longer than 5 minutes are very rare,
with exceptions lasting over an hour. We have detected enhancements over
the total solar irradiance, without exceeding the maximum values
reported for other sites, with more favorable CE conditions.

@@

For the detection of CE cases we established a baseline of irradiance
above which we characterized each data point as an enhancement event and
calculated the over irradiance (OIR). The OIR is defined as the
irradiance difference of the measured one-minute \(\text{GHI}_i\) from
the \(\text{GHI}_\text{ref}\) corresponding to cloud-free atmosphere.
First, we tested two simple approaches for the determination of
\(\text{GHI}_\text{ref}\): The Haurwitz's model \citep{Haurwitz1945},
which is a simple clear sky model and was already adjusted and applied
to our data \citep{Natsis2023}, and the total solar irradiance (TSI) at
the top of the atmosphere, adjusted for the Sun-Earth distance. We have
tested both methods by using an appropriate relative threshold and/or an
additional constant offset. The initial results showed that we can
detect a big portion of the CE events. However, by inspecting the daily
plots of irradiance it became evident that changes in atmospheric
conditions introduced numerous false positive and false negative
results. The main reason for these discrepancies is the variability of
the effects of aerosols and water vapor which were not taken into
account in the two simple methods. To produce a more representative
reference we included the effects of these factors through variables
using a radiative transfer model (RTM). The applied methodology is
discussed in section\nobreakspace{}\ref{rtmcs}..

@@

The above clear sky reference values are based on the climatological AOD
and WC, hence they cannot describe accurately the long term variation of
\(\text{GHI}_\text{ref}\) due to changes in the two atmospheric
constituents, mainly AOD. As reported by \citet{Natsis2023}, there is a
long term brightening effect in the GHI data of Thessaloniki for the
period 1993 -- 2023, which for clear-sky data was attributed to
long-term changes in aerosol effects. Therefore, an adjustment of the
\(\text{GHI}_\text{ref}\) used in this study is necessary. As AERONET
data start only in 2003, we used for the period 1993 -- 2005 estimates
of changes in AOD at \(340\,\text{nm}\) derived from a collocated Brewer
spectrophotometer \citep{Kazadzis2007} to calculate the trend in
\(\text{GHI}_\text{ref}\) during this period.

@@

Following the application of the above discussed methodology to the
entire dataset (6.1 million of one-minute GHI measurements),
\(1.76\,\%\) were identified as CE events and \(0.036\,\%\) as ECE
events. The highest recorded GHI due to CE was
\(1416.6\,\text{W}/\text{m}^2\) on 24~May 2007, corresponding to OIR of
\(345.9\,\text{W}/\text{m}^2\). The stronger CE event of \(49.7\,\%\)
above the clear sky threshold was observed on 28~October 2016. In the
following sections we are discussing the long-term trends and
variability of the CE events as well as of the corresponding OIR and
excess irradiation.

@@

Another interesting aspect of the CE events is their distribution within
the year. Figure\nobreakspace{}\ref{fig:relative-month-occurrences}
shows the monthly box and whisker plot of the CE number of occurrence
normalized with the highest median value that occurs in June, depicting
a clear seasonal cycle. Although CE events are present throughout the
year, the most active period is during May and June. During winter
(December -- February), the number of CE cases is about \(25\,\%\) of
the maximum, while in the intermediate months the number of occurrences
gradually ramps between the maximum and minimum. This seasonality is a
combined effect of different factors, among them the types of clouds,
their frequency of occurrence the seasonally varying relative position
of the sun, as well as the local landscape characteristics that may
influence the formation of clouds. Unfortunately, lack of detailed data
on cloud formation, type and location is not allowing further analysis.
The interannual variability of the monthly CE events is quite high as
manifested by the large monthly extremes, especially in the summer.

@@

\caption{Seasonal variability of the number of CE events normalized to the maximum occurring in June, in the form of a box and whisker plot. The monthly values have been also normalized to the relative abundance of valid GHI observations. The box contains the data between the lower $25\,\%$ and the upper $75\,\%$ percentiles. The thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values and the solid circles are outliers.}\label{fig:relative-month-occurrences}
\end{figure}

The distribution of the number of CE events as a function of OIR is
shown in Figure\nobreakspace{}\ref{fig:ovir-distribution}. Apparently,
there is an inverse relation between the frequency of CE events and OIR
with an exponential-like decline. This is expected, as the stronger the
CE events become, the rarer are the conditions favoring the occurrence
of CE events. For the majority (over \(62\,\%\)) of the CE events the
OIR is below the long term average of \(39.9\,\text{W}/\text{m}^2\),
while about \(8.1\,\%\) of the events correspond to OIR larger than
\(100\,\text{W}/\text{m}^2\) and up to highest value of
\(412.4\,\text{W}/\text{m}^2\). This distribution is indicative of the
magnitude and the probability of the expected CE events over
Thessaloniki. Similar distribution of CE events, albeit with larger OIR
values, has been reported by \citet{Vamvakas2020} for the city of
Patras. This site is located \(2.5^\circ\) south of Thessaloniki and is
exposed to air masses coming mainly from the eastern Mediterranean
resulting in different cloud patterns, that may affect the
characteristics and magnitude of the CE events.

@@

The relation between the duration of the CE groups and the excess
irradiation of the groups has also been investigated
(Figure\nobreakspace{}\ref{fig:group-2d}). Evidently, evens of high
excess irradiation have small duration and vice versa. The vast majority
of grouped events are associated with small excess irradiation (e.g.
\(<5\,\text{kJ}/\text{m}^2\)) and small duration
(e.g.~\(<5\,\text{min}\)) while groups with strong excess irradiation
and long duration are rare. Similar results on this relation, have been
reported by \citet{Zhang2018}, on a study using a far higher sampling
rate than ours.

@@

By creating a clear sky approximation of the GHI, which represents the
long- and short-term variation of the expected clear sky GHI, we were
able to identify cases of CE events. After analyzing the CE cases, we
found an increase of \(45.6\pm 21.9\,\text{cases}/\text{year}\), with
the mean annual total energy of the CE events increasing with a rate of
\(116.9\pm 67.8\,\text{kJ}/\text{year}\). The most active season of CE
events over Thessaloniki is concentrated on early summer, on May and
June.
